---
title: 3. calculations
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. calculations}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include=FALSE}
require(rintcal)
```

The rintcal package also provides some functions related to radiocarbon calibration. First there are two functions to calculate radiocarbon ages from pMC values (in this case of a postbomb date):

```{r}
pMC.age(150, 1)
```

and the other way round:

```{r}
age.pMC(-2300, 40)
```

We can also obtain the calibrated probability distributions from radiocarbon dates, e.g. one of 130 +- 20 C14 BP:

```{r, fig.width=4, fig.asp=.8}
calib.130 <- caldist(130, 20, BCAD=TRUE)
plot(calib.130, type="l")
```

For reporting purposes, calibrated dates are often reduced to their 95% highest posterior density (hpd) ranges (please report all, not just your favourite one!):

```{r}
hpd(calib.130)
```

Additionally, calibrated dates are often reduced to single point estimates. Note however how poor representations they are of the entire calibrated distribution!

```{r, fig.width=4, fig.asp=.8}
calib.2450 <- caldist(2450, 20)
plot(calib.2450, type="l")
points.2450 <- point.estimates(calib.2450)
points.2450
abline(v=points.2450, col=1:4, lty=2)
```

Want a plot of the radiocarbon and calibrated distributions, together with their hpd ranges?
```{r, fig.width=5, fig.asp=1}
calibrate(130,20)
```

You can also draw one or more calibrated distributions:

```{r, fig.width=4, fig.asp=1}
set.seed(123)
dates <- sort(runif(5, 100, 1000))
errors <- .05*dates
depths <- 0:4
labels <- c("my", "very", "own", "simulated", "dates")
draw.dates(dates, errors, depths, BCAD=TRUE, labels=labels, age.lim=range(dates, 600, 1700))
```

or add them to an existing plot:

```{r, fig.width=4, fig.asp=1}
plot(0, type="n", xlim=c(600, 1700), ylim=c(5,-1), xlab="AD", ylab="dates")
draw.dates(dates, errors, depths, BCAD=TRUE, add=TRUE, labels=labels)
```



***

<br>
[<- prev (plots)](./plots.html)
<br>

