---
title: "IntCal - calculations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{calculations}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include=FALSE}
require(IntCal)
```

The IntCal package also provides some functions related to radiocarbon calibration. First there are two functions to calculate radiocarbon ages from pMC values (in this case of a postbomb date):

```{r}
pMC.age(150, 1)
```

and the other way round:

```{r}
age.pMC(-2300, 40)
```

We can also obtain the calibrated probability distributions from radiocarbon dates, e.g. one of 130 +- 20 C14 BP:

```{r, fig.width=4, fig.asp=.8}
calib.130 <- caldist(130, 20, BCAD=TRUE)
plot(calib.130, type="l")
```

For reporting purposes, calibrated dates are often reduced to their 95% highest posterior density (hpd) ranges (please report all, not just your favourite one!):

```{r}
hpd(calib.130)
```

Want a plot of the radiocarbon and calibrated dates, together with their hpd ranges?

```{r, fig.width=5, fig.asp=1}
calibrate(130,20)
```


***

<br>
[<- prev (plots)](./plots.html)
<br>

