---
title: 3. calculations
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. calculations}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include=FALSE}
require(rintcal)
```

The rintcal package also provides some functions related to radiocarbon calibration. First there are two functions to calculate radiocarbon ages from pMC values (in this case of a postbomb date):

```{r}
pMC.age(150, 1)
```

and the other way round:

```{r}
age.pMC(-2300, 40)
```

The same for calculations in the F14C realm:

```{r}
F14C.age(.150, .01)
```

and the other way round:

```{r}
age.F14C(-2300, 40)
```

To transfer D14C (deviations in atmospheric 14C concentration at t cal BP) to F14C, and the other way around:

```{r}
F14C.D14C(0.8, t=1000)
D14C.F14C(-97.2, 1000)
```

These functions can be used to investigate Delta14C over time:

```{r, fig.width=4, fig.asp=.8}
cc <- ccurve()
cc.Fmin <- age.F14C(cc[,2]+cc[,3])
cc.Fmax <- age.F14C(cc[,2]-cc[,3])
cc.D14Cmin <- F14C.D14C(cc.Fmin, cc[,1])
cc.D14Cmax <- F14C.D14C(cc.Fmax, cc[,1])
par(mar=c(4,3,1,3), bty="l")
plot(cc[,1]/1e3, cc.D14Cmax, type="l", xlab="kcal BP", ylab="")
mtext(expression(paste(Delta, ""^{14}, "C")), 2, 1.7)
lines(cc[,1]/1e3, cc.D14Cmin)
par(new=TRUE)
plot(cc[,1]/1e3, (cc[,2]+cc[,3])/1e3, type="l", xaxt="n", yaxt="n", col=4, xlab="", ylab="")
lines(cc[,1]/1e3, (cc[,2]-cc[,3])/1e3, col=4)
axis(4, col=4, col.axis=4)
mtext("14C kBP", 4, 1.7, col=4)
```

The above functions can be used to calculate the effect of contamination on radiocarbon ages, e.g. what age would be observed if material with a 'true' radiocarbon age of 5000 +- 20 14C BP would be contaminated with 1% of modern carbon (F14C=1)?

```{r}
contaminate(5000, 20, .01, 1)
```

The effect of different levels of contamination can also be visualised:

```{r, fig.width=4, fig.asp=.8}
real.14C <- seq(0, 50e3, length=200)
contam <- seq(0, .1, length=101) # 0 to 10% contamination
contam.col <- rainbow(length(contam))
plot(0, type="n", xlim=c(0, 55e3), xlab="real 14C age", ylim=range(real.14C), ylab="observed 14C age")
for(i in 1:length(contam))
  lines(real.14C, contamination(real.14C, c(), contam[i], 1, decimals=5), col=contam.col[i])
contam.legend <- seq(0, .1, length=6)
contam.col <- rainbow(length(contam.legend))
text(52e3, contaminate(50e3, c(), contam.legend, 1), labels=contam.legend, col=contam.col, cex=.7)

```


Now on to calibration. We can obtain the calibrated probability distributions from radiocarbon dates, e.g. one of 130 +- 20 C14 BP:

```{r, fig.width=4, fig.asp=.8}
calib.130 <- caldist(130, 20, BCAD=TRUE)
plot(calib.130, type="l")
```

For reporting purposes, calibrated dates are often reduced to their 95% highest posterior density (hpd) ranges (please report all, not just your favourite one!):

```{r}
hpd(calib.130)
```

Additionally, calibrated dates are often reduced to single point estimates. Note however how poor representations they are of the entire calibrated distribution!

```{r, fig.width=4, fig.asp=.8}
calib.2450 <- caldist(2450, 20)
plot(calib.2450, type="l")
points.2450 <- point.estimates(calib.2450)
points.2450
abline(v=points.2450, col=1:4, lty=2)
```

Want a plot of the radiocarbon and calibrated distributions, together with their hpd ranges?
```{r, fig.width=5, fig.asp=1}
calibrate(130,20)
```

You can also draw one or more calibrated distributions:

```{r, fig.width=4, fig.asp=1}
set.seed(123)
dates <- sort(runif(5, 100, 1000))
errors <- .05*dates
depths <- 0:4
labels <- c("my", "very", "own", "simulated", "dates")
draw.dates(dates, errors, depths, BCAD=TRUE, labels=labels, age.lim=range(dates, 600, 1700))
```

or add them to an existing plot:

```{r, fig.width=4, fig.asp=1}
plot(0, type="n", xlim=c(600, 1700), ylim=c(5,-1), xlab="AD", ylab="dates")
draw.dates(dates, errors, depths, BCAD=TRUE, add=TRUE, labels=labels)
```



***

<br>
[<- prev (plots)](./plots.html)
<br>

