---
title: IntCal - introduction
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

The international IntCal group provides ratified radiocarbon calibration curves such as IntCal20 (for northern hemisphere terrestrial radiocarbon dates; Reimer et al. 2020[^1]), Marine20 (for marine dates; Heaton et al. 2020[^2]) and SHCal20 (Hogg et al. 2020[^3]). This package provides these curves, as well as previous iterations (IntCal13, Marine13, SHCal13, IntCal09, Marine09, IntCal04, Marine04, SHCal04, IntCal98, Marine98) and postbomb curves (Hua et al. 2013[^4], Levin and Kromer 2004[^5], Santos et al. 2015[^6], Andrews et al. 2016[^7]). 

On first usage of the package, it has to be installed: 

```{r, eval=FALSE}
install.packages('IntCal')
``` 

If you have a recent version of rbacon, rplum or clam installed on your computer, IntCal will probably have been installed as well. Sometimes new versions of these packages appear, so please re-issue the above command regularly to remain up-to-date, or use:

```{r, eval=FALSE}
update.packages()
``` 

To obtain access to the calibration curves, first the package has to be loaded:

```{r}
library(IntCal)
```

Now you can load a calibration curve into the memory, for example the default curve IntCal20, and check the first few entries:

```{r}
ic20 <- ccurve()
head(ic20)
```

The files have three columns: cal BP, the corresponding IntCal C14 BP ages, and the uncertainties (1 standard deviation). 

To see more detail of each IntCal function, place a question-mark before the function name, e.g.:

```{r, eval=FALSE}
?ccurve
```

To get a list of available curves and associated files (and where they can be found):

```{r}
list.ccurves()
```

You can also combine calibration curves, e.g. a 40%/60% mix of Intcal20 and Marine20 with a 100+-20 year offset for the latter. The resulting curve will be saved with the name mixed.14C, in a folder together with the calibration curves. The name of this folder is listed, and it can be changed by specifying the option 'dir'.

```{r}
mix.ccurves(0.4, cc1="IntCal20", cc2="Marine20", offset=c(100, 20))
```

To glue prebomb and postbomb calibration curves into one and store it as a variable in your session (for example, IntCal20 and the NH1 postbomb curve):

```{r, fig.width=4, fig.asp=.8}
glued <- glue.ccurves("IntCal20", "NH1")
plot(glued[1:650,1:2], xlab="cal BP", ylab="C-14 BP", pch=".")
```

***

<br>
[next (plots) ->](./plots.html)
<br>

[^1]: Reimer PJ et al., 2020. The IntCal20 Northern Hemisphere radiocarbon age calibration curve (0–55 cal kBP). Radiocarbon 62, 725-757 <!-- [doi:10.1017/RDC.2020.41](http://dx.doi.org/10.1017/RDC.2020.41){target="_blank"} -->

[^2]: Heaton TJ et al., 2020. Marine20—The Marine Radiocarbon Age Calibration Curve (0–55,000 cal BP). Radiocarbon 62, 779-820 <!--[doi:10.1017/RDC.2020.68](http://dx.doi.org/10.1017/RDC.2020.68){target="_blank"} -->

[^3]: Hogg AG et al., 2020. SHCal20 Southern Hemisphere Calibration, 0–55,000 Years cal BP. Radiocarbon 62, 759-778 <!--[doi:10.1017/RDC.2020.59](http://dx.doi.org/10.1017/RDC.2020.59){target="_blank"} -->

[^4]: Hua Q et al. 2013. Atmospheric Radiocarbon for the Period 1950–2010 Radiocarbon 55, 2059-2072. <!--[doi:10.2458/azu_js_rc.v55i2.16177](http://dx.doi.org/10.2458/azu_js_rc.v55i2.16177){target="_blank"} -->

[^5]: Levin I, Kromer, B,  2004. The Tropospheric 14CO2 Level in Mid-Latitudes of the Northern Hemisphere (1959-2003), Radiocarbon 46, 1261-1272 <!--[doi:10.1017/S0033822200033130](http://dx.doi.org/10.1017/S0033822200033130){target="_blank"} -->

[^6]: Santos GM, Linares R, Lisi CS, Filho MT, 2015. Annual growth rings in a sample of Parana pine (*Araucaria angustifolia*): Toward improving the 14C calibration curve for the Southern Hemisphere. Quaternary Geochronology 25, 96-103 <!--[doi:10.1016/j.quageo.2014.10.004](http://dx.doi.org/10.1016/j.quageo.2014.10.004){target="_blank"} -->

[^7]: Andrews H, Siciliano D, Potts DC, DeMartini EE, Covarrubias S, 2016. Bomb radiocarbon and the Hawaiian Archipelago: Coral, otoliths and seawater. Radiocarbon 58, 531-548 <!--[doi:10.1017/RDC.2016.32](http://dx.doi.org/10.1017/RDC.2016.32){target="_blank"} -->
