---
title: "Introduction to the IntCal package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"intro"}
  %\VignetteEngine{knitr::rmarkdown}
---

The international IntCal group provides ratified radiocarbon calibration curves such as IntCal20 (for northern hemisphere terrestrial radiocarbon dates; Reimer et al. 2020[^1]), Marine20 (for marine dates; Heaton et al. 2020[^2]) and SHCal20 (Hogg et al. 2020[^3]). This package provides these curves, as well as previous iterations (IntCal13, Marine13, SHCal13, IntCal09, Marine09, IntCal04, Marine04, SHCal04, IntCal98, Marine98) and postbomb curves (Hua et al. 2013[^4], Levin and Kromer 2004[^5], Santos et al. 2015[^6], Andrews et al. 2016[^7]). 

On first usage of the package, it has to be installed: 

```{r, eval=FALSE}
install.packages('IntCal')
``` 

If you have a recent version of rbacon, rplum or clam installed on your computer, IntCal will probably have been installed as well. Sometimes new versions of these packages appear, so please re-issue the above command regularly to maintain up-to-date, or use:

```{r, eval=FALSE}
update.packages()
``` 

To obtain access to the calibration curves, first the package has to be loaded:

```{r}
library(IntCal)
```

Now you can load a calibration curve into the memory, for example the default curve IntCal20, and check the first few entries:

```{r}
ic20 <- ccurve()
head(ic20)
```

The files have three columns: cal BP, the corresponding IntCal C14 BP ages, and the uncertainties (1 standard deviation). 

To get a list of available curves and associated files (and where they can be found):

```{r}
list.ccurves()
```

You can also combine calibration curves, e.g. a 40%/60% mix of Intcal20 and Marine20 with a 100+-20 year offset for the latter. The resulting curve will be saved together with the calibration curves, with the name mixed.14C:

```{r}
mix.curves(0.4, cc1="IntCal20", cc2="Marine20", offset=c(100, 20))
mycurve <- ccurve("mixed")
head(mycurve)
```


***

<br>
[next (plots) ->](./plots.html)
<br>

[^1]: Reimer PJ et al., 2020. The IntCal20 Northern Hemisphere radiocarbon age calibration curve (0–55 cal kBP). Radiocarbon 62, 725-757. doi:10.1017/RDC.2020.41

[^2]: Heaton TJ et al., 2020. Marine20—The Marine Radiocarbon Age Calibration Curve (0–55,000 cal BP). Radiocarbon 62, 779-820. doi:10.1017/RDC.2020.68

[^3]: Hogg AG et al., 2020. SHCal20 Southern Hemisphere Calibration, 0–55,000 Years cal BP. Radiocarbon 62, 759-778. doi:10.1017/RDC.2020.59

[^4]: Hua Q et al. 2013. Atmospheric Radiocarbon for the Period 1950–2010. Radiocarbon, 2059-2072. doi:10.2458/azu_js_rc.v55i2.16177

[^5]: Levin I, Kromer, B,  2004. The Tropospheric 14CO2 Level in Mid-Latitudes of the Northern Hemisphere (1959-2003), Radiocarbon 46(3):1261-1272

[^6]: Santos GM, Linares R, Lisi CS, Filho MT, 2015. Annual growth rings in a sample of Parana pine (Araucaria angustifolia): Toward improving the 14C calibration curve for the Southern Hemisphere. Quaternary Geochronology 25:96-103

[^7]: Andrews H, Siciliano D, Potts DC, DeMartini EE, Covarrubias S, 2016. Bomb radiocarbon and the Hawaiian Archipelago: Coral, otoliths and seawater. Radiocarbon 58:531-548 <doi:10.1017/RDC.2016.32>



