---
title: 2. plots
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. plots}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE}
require(rintcal)
```

We can also plot a calibration curve:

```{r, fig.width=4, fig.asp=.8}
draw.ccurve()
```

You can also compare two calibration curves:

```{r, fig.width=4, fig.asp=.8}
draw.ccurve(1600, 2020, BCAD=TRUE, cc2='nh1', add.yaxis=TRUE)
```

Or zoom in to between AD 1600 and 2000 (using the BCAD scale):

```{r, fig.width=4, fig.asp=.8}
draw.ccurve(1600, 1950, BCAD=TRUE)
```

Interesting things happened after 1950, as can be seen by adding a postbomb curve:

```{r, fig.width=4, fig.asp=.8}
draw.ccurve(1600, 2020, BCAD=TRUE, cc2='nh1')
```

The postbomb curve dwarfes the IntCal20 curve, so we could also plot both on separate vertical axes:

```{r, fig.width=4, fig.asp=.8}
draw.ccurve(1600, 2020, BCAD=TRUE, cc2='nh1', add.yaxis=TRUE)
```

We can also visualise the data underlying parts of the IntCal calibration curve, for example from 500 to 0 cal BP:

```{r, fig.width=5, fig.asp=1}
intcal.data(0, 500)
```


***

<br>
[<- prev (intro)](./intro.html) | 
[next (calculations) ->](./calculations.html)
<br>

